The use of lithic assemblages for the definition of short-term occupations in hunter-gatherer prehistory

### 2018-06-06 analysis/paper/paper.Rmd
final revision

In Binford’s seminal framework there are clear definitions of site typology [@binford_willow_1980] that relate to mobility patterns:
“For foragers, I recognized two types of site, the residential base camp and the location. Collectors generate at least three additional types of sites by virtue of the logistical character of their procurement strategies. These I have designated the field camp, the station, and the cache.” [@binford_willow_1980, p. 10].

Another good example of site nomenclature is that of Dillehay et al. [-@dillehay_research_2011] for the foraging-farming transition in the Andes. The authors indicate the presence of a very diverse and expanded site typology [@dillehay_research_2011, pp. 36-40], defining long-term and short-term base and field camps, processing stations, transitory station/workshops, lithic quarries, earthen mounds, horticultural residences with gardens, permanent residences associated with irrigation, agriculture, hillside villages and special activity locales.

These examples start with a basic assumption that “…the greater the number of generic types of functions a site may serve, the greater the number of possible combinations, and hence the greater the range of inter-site variability…” [@binford_willow_1980, p. 12]. However, as Binford also argued, there is considerable variability in the duration of each stay in each site and the type of mobility pattern present [@binford_willow_1980], and thus each site may have been used for different functions and durations [@binford_archaeology_1982], with a direct impact on the archaeological visibility of sites and their respective duration cycle. Barton and colleagues [@riel-salvatore_late_2004; @barton_formation_2014; @clark_lithics_2017], following Binford’s perspective, have argued for a continuum in the use of space and time between residential (Foragers) and logistical (Collectors) patterns, that can be measured in the archaeological record using the relationship between artifact volumetric density and the frequency of retouched tools within each assemblage – named by the authors as the Whole Assemblage Behavioral Index (WABI). The index is expected to show a negative correlation between both variables, reflecting accumulated artifacts deriving from primarily curated to primarily expedient artifact use.  Although the authors emphasize that “the terms “expedient” and “curated” do not reflect individual site-occupation events” [@riel-salvatore_new_2007, p. 62]. They also assume that expedient assemblages often accumulate at more intensively occupied sites, while curated assemblages more usually derive from short-term occupations.

@@

ggplot(dataset, aes(x=dataset$LithicDensity, y=dataset$RetouchFreqPerc)) +
  geom_point(shape=16, size=3) +
  labs(title = "Whole Assemblage Behavioral Index", x = "Lithic Volumetric Density (Artifacts/m3) (Log10)", y = "Frequency of Retouched Pieces (Log10)") +
  geom_text_repel(label=dataset$Sites) +
  geom_smooth(method="lm", level = 0.80, formula = y~x) +
  scale_x_continuous(trans='log10' ) +
  scale_y_continuous(trans='log10',  limits = c(NA,100))

@@

Similarly, sites which were occupied as field camps, using Binford’s terminology that define these as “temporary operational center[s] for task groups […], where a task group sleeps, eats, and otherwise maintains itself while away from the residential base” [@binford_willow_1980, p. 10], can be associated with both expedient or curated assemblages, depending on a complex set of particularities, of which availability of usable raw materials is a central one. The differences introduced by Kuhn’s [-@kuhn_mousterian_1995] concepts of “provisioning of places” and “provisioning of individuals” are fundamental in this regard, distinguishing between strategies in which groups supply themselves from immediately available resources and make tools on the spot, or they anticipate their needs by transporting raw material blocks or already produced tools [@porraz_middle_2009]. This seems to be the case with the VB Shelter B context. Although occupying a fairly small area of the site, a very large diversity of chert raw materials [@pereira_territory_2016], low presence of cores and blanks, and high frequency of knapping residues [@cascalheira_influencia_2013; @cascalheira_tecnologia_2010], indicate that this context was probably used multiple times as a temporary field camp, where the complex combination of in situ knapping and retooling activities, resulted in a rather typical expedient assemblage. Thus, expedient behaviors are not conditioned by the amount of time spent in a site, but as detailed by Nelson [-@nelson_study_1991] on the location of activities close to raw material sources (or stockpiling), the lack of time stress in tool manufacture, and the regular use of sites that allow people to take advantage of abundant, predictable, resources.

@@

Also problematic in our opinion, is the dichotomy of “curated technologies = residential mobility” vs “expedient technologies = logistical mobility” as presented by, for example, Riel-Salvatore [-@riel-salvatore_niche_2010]. The case of VB Shelter 2, previously classified as a Gravettian point cache [@bicho_early_2016], and whose location both in the WABI plot and in the PCA biplots revealed a highly curated assemblage, is particularly relevant in this context. Although rare in the Eurasian Paleolithic archaeological record [but see e.g. @aubry_les_2003; @tabarev_paleolithic_2013 for other examples], based on the ethnographic record cache contexts are thought to be “common components of a logistical strategy in that successful procurement of resources by relatively small groups for relatively large groups generally means large bulk” [@binford_willow_1980, p. 12]. It is true, though, that in the context of the classification of occupation duration, lithic caches offer an additional classificatory problem, since site use of these contexts can be classified as either a very short occupation (when the cache is placed at the specified location and picked up when needed), or as long-term occupations if considering the total amount of time that the cache is in use. Either way, a context like VB Shelter 2 is unlikely to be associated with a residential mobility strategy, possibly representing the shorter and more curated type of assemblages of a logistical system. From a different perspective, accepting that significant time investment (both in tool manufacture and use life) is an essential component in characterizing reliable assemblages [sensu @bleed_optimal_1986], frequently associated with high levels of curation, then the VB Shelter 2 does not seem to fit at all in that category. The set of small retouched and unretouched bladelets more plausibly fit a maintainable type of system, that in the WABI calculation are associated with residential settlement systems. This particular contradiction mirrors the strong debate carried over the years about the factors explaining the curated/expedient dichotomy [see @vaquero_searching_2017 for a comprehensive review], and their repercussive implications for settlement interpretation.

>>>>

In Binford’s seminal framework there are clear definitions of site typology [@binford_willow_1980] that relate to mobility patterns:
“For foragers, I recognized two types of site, the residential base camp and the location. Collectors generate at least three additional types of sites by virtue of the logistical character of their procurement strategies. These I have designated the field camp, the station, and the cache.” [@binford_willow_1980: 10].

Another good example of site nomenclature is that of Dillehay et al. [-@dillehay_research_2011] for the foraging-farming transition in the Andes. The authors indicate the presence of a very diverse and expanded site typology [@dillehay_research_2011: 36-40], defining long-term and short-term base and field camps, processing stations, transitory station/workshops, lithic quarries, earthen mounds, horticultural residences with gardens, permanent residences associated with irrigation, agriculture, hillside villages and special activity locales.

These examples start with a basic assumption that “…the greater the number of generic types of functions a site may serve, the greater the number of possible combinations, and hence the greater the range of inter-site variability…” [@binford_willow_1980: 12]. However, as Binford also argued, there is considerable variability in the duration of each stay in each site and the type of mobility pattern present [@binford_willow_1980], and thus each site may have been used for different functions and durations [@binford_archaeology_1982], with a direct impact on the archaeological visibility of sites and their respective duration cycle. Barton and colleagues [@riel-salvatore_late_2004; @barton_formation_2014; @clark_lithics_2017], following Binford’s perspective, have argued for a continuum in the use of space and time between residential (Foragers) and logistical (Collectors) patterns, that can be measured in the archaeological record using the relationship between artifact volumetric density and the frequency of retouched tools within each assemblage – named by the authors as the Whole Assemblage Behavioral Index (WABI). The index is expected to show a negative correlation between both variables, reflecting accumulated artifacts deriving from primarily curated to primarily expedient artifact use.  Although the authors emphasize that “the terms “expedient” and “curated” do not reflect individual site-occupation events” [@riel-salvatore_new_2007: 62]. They also assume that expedient assemblages often accumulate at more intensively occupied sites, while curated assemblages more usually derive from short-term occupations.

@@

ggplot(dataset, aes(x=dataset$LithicDensity, y=dataset$RetouchFreqPerc)) +
  geom_point(shape=16, size=3) +
  labs(x = "Lithic Volumetric Density (Artifacts/m3) (Log10)", y = "Frequency of Retouched Pieces (Log10)") +
  geom_text_repel(label=dataset$Sites) +
  geom_smooth(method="lm", level = 0.80, formula = y~x) +
  scale_x_continuous(trans='log10' ) +
  scale_y_continuous(trans='log10',  limits = c(NA,100))

@@

Similarly, sites which were occupied as field camps, using Binford’s terminology that define these as “temporary operational center[s] for task groups […], where a task group sleeps, eats, and otherwise maintains itself while away from the residential base” [@binford_willow_1980: 10], can be associated with both expedient or curated assemblages, depending on a complex set of particularities, of which availability of usable raw materials is a central one. The differences introduced by Kuhn’s [-@kuhn_mousterian_1995] concepts of “provisioning of places” and “provisioning of individuals” are fundamental in this regard, distinguishing between strategies in which groups supply themselves from immediately available resources and make tools on the spot, or they anticipate their needs by transporting raw material blocks or already produced tools [@porraz_middle_2009]. This seems to be the case with the VB Shelter B context. Although occupying a fairly small area of the site, a very large diversity of chert raw materials [@pereira_territory_2016], low presence of cores and blanks, and high frequency of knapping residues [@cascalheira_influencia_2013; @cascalheira_tecnologia_2010], indicate that this context was probably used multiple times as a temporary field camp, where the complex combination of in situ knapping and retooling activities, resulted in a rather typical expedient assemblage. Thus, expedient behaviors are not conditioned by the amount of time spent in a site, but as detailed by Nelson [-@nelson_study_1991] on the location of activities close to raw material sources (or stockpiling), the lack of time stress in tool manufacture, and the regular use of sites that allow people to take advantage of abundant, predictable, resources.

@@

Also problematic in our opinion, is the dichotomy of “curated technologies = residential mobility” vs “expedient technologies = logistical mobility” as presented by, for example, Riel-Salvatore [-@riel-salvatore_niche_2010]. The case of VB Shelter 2, previously classified as a Gravettian point cache [@bicho_early_2016], and whose location both in the WABI plot and in the PCA biplots revealed a highly curated assemblage, is particularly relevant in this context. Although rare in the Eurasian Paleolithic archaeological record [but see e.g. @aubry_les_2003; @tabarev_paleolithic_2013 for other examples], based on the ethnographic record cache contexts are thought to be “common components of a logistical strategy in that successful procurement of resources by relatively small groups for relatively large groups generally means large bulk” [@binford_willow_1980: 12]. It is true, though, that in the context of the classification of occupation duration, lithic caches offer an additional classificatory problem, since site use of these contexts can be classified as either a very short occupation (when the cache is placed at the specified location and picked up when needed), or as long-term occupations if considering the total amount of time that the cache is in use. Either way, a context like VB Shelter 2 is unlikely to be associated with a residential mobility strategy, possibly representing the shorter and more curated type of assemblages of a logistical system. From a different perspective, accepting that significant time investment (both in tool manufacture and use life) is an essential component in characterizing reliable assemblages [sensu @bleed_optimal_1986], frequently associated with high levels of curation, then the VB Shelter 2 does not seem to fit at all in that category. The set of small retouched and unretouched bladelets more plausibly fit a maintainable type of system, that in the WABI calculation are associated with residential settlement systems. This particular contradiction mirrors the strong debate carried over the years about the factors explaining the curated/expedient dichotomy [see @vaquero_searching_2017 for a comprehensive review], and their repercussive implications for settlement interpretation.
